---
title: Exposure of CD3 T cells to tenofovir
output:
  md_document:
    variant: markdown_github
---

```{r,echo=FALSE,message=FALSE, warning=FALSE}
library(dplyr)
library(lumi)
library(limma)

file <- "J:\\MacLabUsers\\HLADIK SHARED\\Projects\\Herpes study\\Herpes-Project-1\\vaginal_explant_Illumina\\2015_11_10/GenomeStudioProject/SeanHughes_HumanHT12v4_151112/2015.11.12smhughesFinalReport.txt"

rawData <- lumiR(file, detectionTh = 0.05, convertNuID = FALSE, 
   checkDupId = FALSE, columnNameGrepPattern = list(exprs = 'AVG_SIGNAL',
      se.exprs = 'BEAD_STDERR', detection = 'Detection Pval',
      beadNum = 'Avg_NBEADS'),
   annotationColumn = c('ILMN_GENE', 'ENTREZ_GENE_ID', 'GI',
      'ACCESSION', 'SYMBOL', 'PROBE_ID', 'PROBE_START', 'PROBE_SEQUENCE',
      'CHROMOSOME', 'PROBE_CHR_ORIENTATION', 'PROBE_COORDINATES'))

rawData <- rawData[ , 1:5]
```


```{r,echo=FALSE,message=FALSE, warning=FALSE}
# Sample 6 failed, so only have: 
# VEU1567 1600 uM Tenofovir d7 CD3
# VEU1567 0 uM Tenofovir d7 CD3
# VEU3097 1600 uM Tenofovir d7 CD3
# VEU3097 0 uM Tenofovir d7 CD3
# DL2409 1600 uM Tenofovir d7 CD3

# Create the phenotypic data information --------------------------------

pData <- data.frame(Donor = c("A", "A", "B", "B", "C"), 
   Concentration = c(1600, 0, 1600, 0, 1600))

#create metadata df
metadata <- data.frame(
   labelDescription = c("Blood donor", "Concentration of tenofovir"),
   row.names = c("Donor", "Concentration"))

#combine metadata and pdata into an annotated df
adf <- new("AnnotatedDataFrame",data = pData, varMetadata = metadata)

#create experiment data
experimentData <- new("MIAME", name="Sean Hughes", lab = "Florian Hladik Lab",
   title = "CD3 1.6 mM tenofovir exposure")

#make a lumiBatch that contains both the raw data we got from
#shared resources AND the phenoData, metadata and experiment data
#that I created

annotatedData <- new("LumiBatch", exprs=exprs(rawData), phenoData = adf,
                  experimentData = experimentData,
                  se.exprs = se.exprs(rawData),
                  detection = detection(rawData))

```

 SOME PLOTS OF NON NORMALIZED DATA:
 density plot, cdf plot
```{r,echo=FALSE,message=FALSE, warning=FALSE}
#density plot
density(annotatedData)#number of probes for each sample that occur
#at a certain log2 intensity

#CDF plot: cumulative probability of having <= a certain log2 intensity
plotCDF(annotatedData)

#sample relations
plot(annotatedData, what='sampleRelation',method="mds")

#boxplot
boxplot(annotatedData)

###################BACKGROUND CORRECTION ####################
#the data we got from the core had no background correction so I will do it here

QNTB.complete.RAW.lumi <- annotatedData %>% 
   lumiB() %>% 
   lumiT() %>% 
   lumiN(method = "rsn") %>% 
   lumiQ(detectionTh = 0.05)

```
PLOTS OF NORMALIZED DATA 

```{r,echo=FALSE,message=FALSE, warning=FALSE}
plot(QNTB.complete.RAW.lumi)

plot(QNTB.complete.RAW.lumi, what='sampleRelation', method="mds", 
   dimension = c(2, 1), col = pData$Donor)

boxplot(QNTB.complete.RAW.lumi)

```
FILTERING PROBES BASED ON DETECTION 
 Limma suggests to keep probes that are expressed above bg on 
at least n arrays where n is smallest number of replicates assigned
to any of the treatment combinations.

```{r,echo=FALSE,message=FALSE, warning=FALSE}
# this is how the detection filtering works
# Mark the detection pvalues (there is one per probe per sample) with
# a 1 if <0.05 or a 0 if >0.05 using 
# (detection(QNTB.complete.RAW.lumi)<0.05)

# add up the 0's and 1's across each row (i.e. for all the samples)
# using rowSums. Now you have the number of detection p values <0.05 for 
# each probe for all the samples (max possible = total samples)

# now tell me which probes have a rowSum of >=4 (probes will be marked as
# true or false if they do or do not have >=4 rowSum)
            
detectedProbes <- rowSums(detection(QNTB.complete.RAW.lumi)<0.05)>=4

#now extract just those probes that are TRUE from the lumibatch

expressedProbes.lumi <-QNTB.complete.RAW.lumi[detectedProbes,]
```

how many probes did we have before and after filtering?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)

dims(expressedProbes.lumi)
```
how many removed?
```{r,echo=FALSE,message=FALSE, warning=FALSE}
dims(QNTB.complete.RAW.lumi)-dims(expressedProbes.lumi)

```
Here's the design matrix for the analysis

```{r,echo=FALSE,message=FALSE, warning=FALSE}
############### TARGETS AND DESIGN MATRIX ###################
# see section 9.4.1 and 9.4.2 in the limma users guide

targets <- pData(expressedProbes.lumi)
donor <- factor(targets$Donor)
concentration <- factor(targets$Concentration)

design <- model.matrix (~ 0 + concentration + donor)
design
```
Here's the contrasts matrix

```{r,echo=FALSE,message=FALSE, warning=FALSE}
####################### FIT MODEL TO PROBES ###################
fit <- lmFit(expressedProbes.lumi, design = design)

#Now we can make any comparisons
#between the experimental conditions

# If topTable is called and coef has two or more elements,
# then the specified columns will be extracted from fit and
# topTableF called on the result. topTable with coef=NULL is 
# the same as topTableF, unless the fitted model fit has only
# one column.

bothcm <- makeContrasts(
   MaxVsCtrl = concentration1600 - concentration0, 
   levels=design
)
bothcm

#fit the contrasts  
fit2 <- contrasts.fit(fit, bothcm)

#compute diff exprsn
fit2 <- eBayes(fit2)
```

How many probes are up and down regulated for each contrast?

```{r,echo=FALSE,message=FALSE, warning=FALSE}

#method=separate is same as doing topTable for all coefs separately
results <- decideTests(fit2, method="separate", adjust.method="BH",
                      p.value = 0.05, lfc = 0.5)

#turn the results matrix into a data frame and make the
#probeID a real column and remove the rownames

resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing
library(reshape2)

resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

#number of up and down regulated probes based on 
#p.val at least 0.05 and lfc at least 0.5

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))
summary

```


```{r,echo=FALSE,message=FALSE, warning=FALSE}
d <- data.frame(
   x = c(2, 38, 20, 0:40), 
   y = c(10, 10, 5, -(seq(-10, 10, by = 0.5))^2/10), 
   size = c(5, 5, 2, rep(1, 41)))
library(ggplot2)
ggplot(d, aes(x, y)) + geom_point(aes(size = size)) + 
   theme(line = element_blank(), rect = element_blank(), 
      text = element_blank()) + 
   scale_size(guide = FALSE)

```

SessionInfo()

```{r,echo=FALSE,message=FALSE, warning=FALSE}
sessionInfo()
```